Prosper Loan Data Exploration by Samuel McAuliffe
========================================================
<br>

Prosper is an American based peer-to-peer lending platform. The platform is
used to match borrow with investor.When a borrower completes an application for 
funds, their request is reviewed by investors(both institutional and individual)
who are then able to fund up to $35,000 per loan request. Prosper handles the 
servicing of the loan and is responsible for the collection and distribution of 
borrower payments to loan investors. The data encompasses a time frame from
between 2006 and 2013. It should be interesting to see how the GFC affected the
demand for loans around 2007/2008.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(dplyr)
library(knitr)
library(gridExtra)
library(tidyr)
library(lubridate)

# Load the Data
loandf <- read.csv('prosperLoanData.csv')
```

## Examining the structure of the dataset
<br>

The dataset is quite robust, containing 113,937 rows and 81 variables. This is
a considerable amount of variables and inlcudes detailed information about the 
investor, the borrower as well as the loan itself. 

```{r echo=FALSE, message=FALSE, warning=FALSE, structure}
str(loandf)
```

As the dataset has so many variables, for the purpose of this report we shall
focus only on a few select ones which will first be explored using univariate
and summary analysis and then analysed in more detail in the bivariate and
multivariate analysis sections.

Before we explore these variables, I will clean a few of the variables.

```{r  echo=FALSE, message=FALSE, warning=FALSE, Convert_dates}
# Convert loan origination from factor to character
mid <- as.character(loandf$LoanOriginationDate)

#Using lubidate package to replace the original variable with date formatted
loandf$LoanOriginationDate <- ymd_hms(mid)
```

```{r  echo=FALSE, message=FALSE, warning=FALSE, Convert_quarter}
loandf$LoanOriginationQuarter <- as.character(loandf$LoanOriginationQuarter)
loandf <- loandf %>%
         separate (col = LoanOriginationQuarter,
                   into = c("Quarters", "Year"), sep = " ") %>%
         unite(col = LoanOriginationQuarter, Year, Quarters, sep = " ")

loandf$LoanOriginationQuarterF <- factor(loandf$LoanOriginationQuarter)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Convert_credit}

loandf$AverageCreditScore <- (loandf$CreditScoreRangeLower + 
  loandf$CreditScoreRangeUpper) / 2

```

## Converting listing cateogry from numeric to factor.

```{r echo=FALSE, message=FALSE, warning=FALSE, Convert_listing_category}
# Using the excel spreadsheet, convert listing categories to factors.
uses <- c("Not Available", "Debt Consolidation", "Home Improvement", "Business", 
          "Personal Loan", "Student Use", "Auto", "Other", "Baby & Adoption", 
          "Boat", "Cosmetic Procedure", "Engagement Ring", "Green Loans", 
          "Household Expenses", "Large Purchases", "Medical/Dental", 
          "Motorcycle", "RV", "Taxes", "Vacation", "Wedding Loan")

loandf$Categories <- factor(loandf$ListingCategory..numeric., 
                                levels = c(0:20), labels = uses)
```

# Univariate Plots Section
<br>

To get started, we will preform univariate analysis on a number of
the variables to get a feel for the dataset and basic distributions of the
key variables. A short descripton is also provided after each visual. 

## Distribution of Loan Amounts
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, Amount_Summary}
summary(loandf$LoanOriginalAmount)
```

Minimum loan value appears to be $1,000 and the Max is $35,000. The median
loan amount is $6,500. It appears that only 25% of the loans are above $12,000.

```{r echo=FALSE, message=FALSE, warning=FALSE, Loan_Amount_Histogram}
ggplot(aes(x = LoanOriginalAmount), data = loandf) +
  geom_histogram(binwidth = 500, color = 'black', fill = '#57d3ef') +
  scale_x_continuous(limits = c(0, 35000), breaks = seq(0, 35000, 2500))
```

The histogram for loan amounts appears to be a right skewed distribution. There
are however large spikes at prices that are factors of 5 and 10 thousand. The
most noticeable of these spikes are at the 10,000 and 15,000 loan amounts. These
spikes are likely attributed to people simply "rounding up" their required
amount to the nearest $5000. Additionally it could be because there are some
default amounts listed (every $5k) and people just click these rather than enter
their own number. 
Most of the loans are below 15,000 and interestingly our mode appears to be 
$4000. This seems odd given that we have spikes at loan amounts that are
factors of 5 and 10. 

During the sub-prime crisis and resulting GFC, many people used credit to help
fuel their investment/improvement of real estate holdings in the hopes of
flipping it for profit to the next person. As this dataset encompasses the years
of the GFC, I was curious if people who owned a home had different borrowing
patterns than those who didn't.

## Loan Amount Distributions for Home Owners and Non-Home Owners
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, Home_summary}
summary(loandf$IsBorrowerHomeowner)
```

The distribution of home owners to non-home owners is surprisingly fairly even
with 56,459 non-home owners to 57,478 home-owners. 

```{r  echo=FALSE, message=FALSE, warning=FALSE, Amount_vs_home}
ggplot(aes(x = LoanOriginalAmount), data = loandf) +
  geom_histogram(binwidth = 2000, color = 'black', fill = '#57d3ef') +
  scale_x_continuous(limits = c(0, 35000), breaks = seq(0, 35000, 4000)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  facet_wrap(~IsBorrowerHomeowner)
```

The above visual shows the two histograms for those are not home owners and
those that are. It would appear that non-home owners actually had higher counts
in the lower loan amount values (lower than 8k), whereas home-owners had higher 
counts of the higher loans values (8k and above).

## Loan categories
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, Category_summary}
summary(loandf$Categories)
```

There are 21 different categories of reasons for borrowing. The largest of them
is unfortunately "Not Available" and we will exclude this from the plot. The
next largest is debt consolidation with 58,308 loans followed by "Other" at
10,494. 


```{r echo=FALSE, message=FALSE, warning=FALSE,  Counts_of_loan_categories}
# removing the category "Not Available"
ggplot(aes(x = Categories), data = subset(loandf, 
          Categories != "Not Available")) +
  geom_histogram(stat = "count", color = 'black', fill = '#57d3ef') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Unsurprisingly debt consolidation ranked close to 6 times that of the next
highest specific category (home improvement). It was assumed that home 
improvement would also rank high on these loans given the housing bubble that 
was arising in the U.S. at the time. The nondescript "other" category was
actually the second highest overall. 

## Borrower Occupations
<br>

We should also take a quick look at the occupations of borrowers. 

```{r  echo=FALSE, message=FALSE, warning=FALSE, Occupations}
ggplot(aes(x = Occupation), data = loandf) +
  geom_histogram(stat = "count", color = 'black', fill = '#57d3ef') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

The above visual is a little less revealing than I had hoped it be. There is
two obvious spikes, being "Other" and "Professional". They are both very
nondescript and eludes to the possibility that many borrowers are hesitant on
adding their actual profession to their loan request. Instead we shall take a 
look at their employment status.

## Employment Status
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, Employment_summary}
summary(loandf$EmploymentStatus)
```

There are 8 categories of employment. The highest being simply "Employed"
at 67,322. It is unfortunate these individuals aren't categorised into what
form of "Employed" they are. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Employment}
#removed blank employment status category.
ggplot(aes(x = EmploymentStatus), data = subset(loandf, 
          EmploymentStatus != "")) +
  geom_histogram(stat = "count", color = 'black', fill = '#57d3ef') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Employment status tells us some information, however it is still a little
underwhelming. "Employed" has the highest spike, yet it doesn't tell us if these
people are full time, part time, casual or self employed. One has to wonder
whether these sub-par financial assessments are linked to delinquincy. We will
examine this later. Other than the nondescript "Employed" spike, the next
largest group appears to be full time employment, followed by self employed.

As we have been unable to gather any strong signal using the employment status
and occupation variables, we can try examining borrower income levels. 

## Income Range
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, Income_summary}
summary(loandf$IncomeRange)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Income}
ggplot(aes(x = IncomeRange), data = loandf) +
  geom_histogram(stat = "count", color = 'black', fill = '#57d3ef') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_x_discrete(limits = c("Not employed", "$0", "$1-24,999", 
                              "$25,000-49,999", "$50,000-74,999", 
                              "$75,000-99,999", "$100,000+"))
```

I adjusted the above graph to remove the "not displayed" entries and then placed
the x-axis in logical order. It appears that the counts for middle income
earners are the highest. Interestingly those in the upper income brackets have
a higher count than those in the lower. 

## Debt to income Ratio
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, DIR_summary}
summary(loandf$DebtToIncomeRatio)
```

The debt to income ratio ranges from 0 to an enormous 10. The median is 0.22
with the 75% of borrowers falling under 0.32. We will generate a histogram with
an x limit of 1.0 as this represents debt = income situation. The maximum of
10 would represent a party that is either heavily in debt or leveraged to the
hilt on an investment. 

```{r echo=FALSE, message=FALSE, warning=FALSE, DIR}
ggplot(aes(x = DebtToIncomeRatio), data = loandf) +
  geom_bar(stat = "count", binwidth = 0.05, color = 'black', fill = '#57d3ef') +
  scale_x_continuous(limits = c(0, 1.0), breaks = seq(0, 1.0, 0.1))
```

The debt to income ratio histogram is actually quite a nice visual. I limited 
the x-axis to a D:I ratio of 1.0 (debt=income), however there are instances of
borrowers which are beyond this point. The bulk of borrowers have a D:I between
0.05 and 0.4.). A D:I of 0.4 is around the maximum for borrowers seeking a 
a more stable qualified mortgage and both the median and mean fall below this.

## Current Credit Lines
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, CCL_summary}
summary(loandf$CurrentCreditLines)
```

The average is about 10, however the max is quite a large 59. 75% of borrowers
have less than 13 current lines of credit. 


```{r echo=FALSE, message=FALSE, warning=FALSE, CCL}
ggplot(aes(x = CurrentCreditLines), data = loandf) +
  geom_histogram(stat = "count", binwidth = 5, color = 'black', 
                 fill = '#57d3ef') +
  scale_x_continuous(limits = c(0, 40), breaks = seq(0, 40, 5))
```

The histogram of current credit lines is skewed right and centered around 7.5.

## Employment Status Duration
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, EmploymentDuration_summary}
summary(loandf$EmploymentStatusDuration/12)
```

The duration of employment variable is in months so I have divided it by 12 to
convert it to years. We find that the median is about 5.5 with most of the
borrowers having been employed for less than 11.5 years. This would make sense
given that an online p2p loan platform would appeal to younger generations
over older ones. The max is 62.9 years which is quite an outlier. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Employment12}
#Employment status duration converted from months to years
ggplot(aes(x = EmploymentStatusDuration/12), data = loandf) +
  geom_line(stat = "count", binwidth = 1, color = 'black') +
  scale_x_continuous(limits = c(0, 60), breaks = seq(0, 60, 5))
```

As expected employment status duration is skewed right. From the line and 
the summary above we can gather that 50% of borrowers had been employed for less 
than 6 years. The average time was 8 years and the max was a surprising 63.
Another interesting thing is that our first quartile is sitting at 2.2, so 25%
of people have less than 2.2 years employment. I do wonder if this is simply
because of age or possibly because of the inability to find employment.

## Credit Scores
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, PR_summary}
summary(loandf$ProsperRating..numeric.)
```

Our average and median are about 4 which corresponds to a C rating. 

```{r echo=FALSE, message=FALSE, warning=FALSE, PR}
ggplot(aes(x = ProsperRating..numeric.), data = loandf) +
  geom_bar(color = 'black', fill = '#57d3ef') + 
  scale_x_discrete(name = "Prosper Ratings", limits = c(1, 2, 3, 4, 5, 6, 7),
                   labels = c("1" = "HR", "2" = "E", "3" = "D", "4" = "C", 
                              "5" = "B", "6" = "A", "7" = "AA"))
```

The above visual shows the counts of the different loan ratings from high
risk (HR) to safe AA grade. The plot has a fairly normal distribution and the
mean and median are about a C rating. It would be interesting to see the
distribution of these ratings over time.

## Average Credit Rating
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, CS_summary}
summary(loandf$AverageCreditScore)
```

The average credit score is around 690. This is actually better than I had
imagined it would be. The minimum is an abysmal 9.5 and the max is 890. It is
a little concerning that 75% of borrowers are below 730, barely scraping in
to the satisfactory zone. 

```{r echo=FALSE, message=FALSE, warning=FALSE, CS}
ggplot(aes(x = AverageCreditScore), data = loandf) +
  geom_histogram(color = 'black', fill = '#57d3ef', binwidth = 20) +
  scale_x_continuous(limits = c(400, 900), breaks = seq(400, 900, 50))
```

Average credit score has a relatively normal distribution centered around 650-700. 

## Borrower APR
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, APR_summary}
summary(loandf$BorrowerAPR)
```

Median and mean are around the 20 - 22% mark. The lowest is 0.6% and the highest
is 51.2%. The third quartile is around the 28% mark.

```{r echo=FALSE, message=FALSE, warning=FALSE, APR}
ggplot(aes(x = BorrowerAPR), data = loandf) +
  geom_histogram(color = 'black', fill = '#57d3ef', binwidth = 0.02)
```

Relatively normal distribution of APR centered around the 0.20 mark.There is a
a spike at the 0.36 mark. I'm not quite sure what that could be. 

## Credit Score Bins
<br>

We will create a new variable that will hold binned valued for a borrowers
average credit score. These bins are associated average credit scores are as
follows - 

Very Unsatisfactory (<560)
Unsatisfactory (560 - 719)
Satisfactory (720 - 879)
Good (880 - 959)
Excellent (960 - 1000)
```{r echo=FALSE, message=FALSE, warning=FALSE, Scorebins}
#Binning credit scores into factors
loandf <- loandf %>%
  mutate(CreditBins = ifelse(AverageCreditScore < 560, "Very Unsatisfactory",
                      ifelse(AverageCreditScore < 720, "Unsatisfactory",
                      ifelse(AverageCreditScore < 880, "Satisfactory",
                      ifelse(AverageCreditScore < 960, "Good",
                      ifelse(AverageCreditScore < 1000, "Excellent",
                             "Not Available"))))))
loandf$CreditBins <- factor(loandf$CreditBins) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Credbin_summary}
summary(loandf$CreditBins)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Creditbin_uni}
ggplot(aes(x = CreditBins), data = loandf) +
  geom_bar(color = 'black', fill = '#57d3ef')
```

As we can see the Unsatisfactory group has the highest count with over 70k. The
next highest is Satisfactory at 38k. There is only 27 borrowers with a credit
rating over 880. This is actually not surprising as those with sub-par credit
ratings would be drawn to this sort of p2p loan facility. 

# Univariate Analysis Summary
<br>

### What is the structure of your dataset?


Using Univariate analysis we were able to get a feel for the dataset. It is
rather robust at 81 variables. One major issue is that some of the variables
have generalised answers where borrowers have declined to provide information,
or have given a nondescript answer.

### What is/are the main feature(s) of interest in your dataset?


Examining this dataset completely would be far beyond the scope of this project.
Instead I will focus on some of the above variables and their relationship with
each other. I would like to see if there were changes to lending and delinquency
patterns just prior to the GFC. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Described above.

### Did you create any new variables from existing variables in the dataset?


Not yet, just modified some.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Described above.


#Bivariate Plots
<br>

We will now look at our variables in a little more detail and begin to
analyse the relationships between them. As this dataset is all about loans,
we will start by taking a look at loan amounts over time.

##Loan Size Distribution Over Time. 
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, Price_time}
ggplot(aes(x = LoanOriginationDate, y = LoanOriginalAmount), data = loandf) +
  geom_point(alpha = 0.05, position = position_jitter(h = 0), color = 'black') +
    geom_line(stat = 'summary', fun.y = mean, color = "red")
```

While the above visual is heavily overplotted a few major things stand out. 

It appears that between late 2008 and mid 2009 very few new loans seem to be 
created. After some research online I found that the reason for this is that in 
October 2008 Propser decided to stop all new lending on it's site because of 
scrutiny by the SEC. Prosper then announced in late April 2009 that it reopen 
it's lending platform in california (seen as the faint spike in loans at the 
center of the gap in the above visual). Finally in July it began brokering in 
46 states again which is where the gap in the visual above ends.

Another thing that stands out is the emphasised horizontal price lines at 
the $5k intervals. It is likely that people simply rounded up to the nearest
$5k when selecting their desired loan amount.

The mean line in red also shows a rise during the peak of the sub-prime mortgage
crisis. Is it possible that some of the loans through Prosper were also
sub-prime?

One final observation is that at about the end of the first quarter of 2013 it
appears that Prosper may have increased their maximum loan limit as we see
the emergence of loans that exceed $25k.

##Loan amounts for each loan category. 
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, Category_amount}
ggplot(aes(x = Categories, y = LoanOriginalAmount), 
       data = subset(loandf, Categories != "Not Available")) +
  geom_boxplot(lwd = 0.40, outlier.color = I("#C70039"), outlier.shape = 3) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_cartesian(ylim = c(0, 40000))
```

We can see that debt consolidation and baby/adoption categories have the largest
loan amounts. This result for debt consolidation was expected, however the
baby/adoption category having loans this large was a surprise. Other categories
that have high loan amounts include business, green loans and wedding loans. 

## Delinquencies per Quarter
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, DelQuart}
ggplot(aes(x = LoanOriginationQuarter, y = CurrentDelinquencies), 
       data = loandf) +
  geom_bar(stat = "identity", fill = "#57d3ef") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

The above bar graph is quite interesting. We see a big spike in current
delinquencies in the second half of 2006 and first half of 2007. The GFC is
commonly believed to have begun in July 2007. It was set off by a speculative
credit fueled housing bubble. Could this spike be an indicator that some
families were running out of money and were unable to service their loans?

## Average Quarterly Credit Score
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, ScoreQuart}
ggplot(aes(x = LoanOriginationQuarter, y = AverageCreditScore), 
       data = loandf) +
  geom_boxplot(lwd = 0.45, outlier.color = I("#C70039"), outlier.shape = 3) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_cartesian(ylim = c(350, 950))
```

The above plot shows that pre GFC there were quite a number of borrowers with
poor credit scores. Anything below about 650 can be considered a less than ideal
credit score. In fact the median for these years appears to be less than the
standard "good score" of 700. It appears that when the company re-opened to all
states in 2009, they also implemented some new rules for minimum credit scores.
It can be see that beginning in Q3 2009, the median is a little over 700 and the
lower extreme does not dip below 600. Could this possibly show that credit
providers did make some changes following the GFC?

## Credit lines by Income Range
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, CCLIncome}
ggplot(aes(x = IncomeRange, y = CurrentCreditLines), data = loandf) +
  geom_boxplot(lwd = 0.50, outlier.color = I("#C70039"), outlier.shape = 3) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_x_discrete(limits = c("Not employed", "$0", "$1-24,999", 
                              "$25,000-49,999", "$50,000-74,999", 
                              "$75,000-99,999", "$100,000+"))


```

Median current credit lines appears to increase as income rises. This trend does
not hold for those with an income of 0. The IQR for those with $0 income is
the largest of all the groups.

## Employment Length and Loan Status
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, EmployCredit}
ggplot() + geom_histogram(data=loandf, aes(x=EmploymentStatusDuration/12, 
                                        fill=CreditBins),binwidth=1) +
    scale_fill_manual(values = c("#1976D2","#388E3C", "#F57C00", "#D32F2F"))

```

We can see that the portion of borrowers with very unsatisfactory credit scores
are fairly concentrated in the lower employment duration area. Most are under 5
years employment duration. Could this be because many of these individuals are
jumping between part-time jobs and are sometimes unable to service their credit,
leading to the bad credit scores?

Unsatisfactory level retains a higher count than satisfactory for most of the 
graph and this is to be expected given this service would likely attract those 
who are unable to get a suitable interest rate when applying for a loan at a 
financial institution such as a bank. 

## Borrower APR to Credit Rating
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, CreditBorrow}
ggplot(aes(x = CreditBins, y = BorrowerRate * 100),
      data = subset(loandf, !is.na(BorrowerRate))) +
  geom_boxplot(lwd = 0.50, outlier.color = I("#C70039"), outlier.shape = 3)
```

As we can see in the above visual, the borrower rate is somewhat negatively 
correlated with credit score. The median for "good" credit scores is less than 
10% and the median for "very unsatisfactory" appears to be a little over 26%. 
The NA values have a median of about 15, placing them in the risk area somewhere
between satisfactory and unsatisfactory. 

# Bivariate Analysis
<br>

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Described above.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Described above.

### What was the strongest relationship you found?

Strongest relationship would be the borrower rate and the credit score (bins)

# Multivariate Plots Section
<br>

## Quarterly Delinquencies by Income Range
<br>

```{r  echo=FALSE, message=FALSE, warning=FALSE, QuartDelIncome}
ggplot(aes(x = LoanOriginationQuarter, y = CurrentDelinquencies, 
           fill = IncomeRange), data = loandf) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_fill_manual(values = c("#D81B60","#FBC02D", "#388E3C", "#F57C00", 
                               "#7b1FA2", "#D32F2F", "#1976D2", "#212121"))
```

The above graph shows the distribution of current delinquencies for each income
range across time. I had hoped it would reveal some information regarding
the income range of delinquent borrowers in the lead up to the GFC, however it
appears that the "Income range" application question was only introduced
in the first few months of 2007. We can still see that most of the current
delinquencies are from those in the 25 - 50k income range. Instead we might try
using bin values for credit rating. 

## Quarterly Delinquencies by Credit Bin
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, QuartDelCredit}
ggplot(aes(x = LoanOriginationQuarter, y = CurrentDelinquencies, 
           fill = CreditBins), data = loandf) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_fill_manual(values = c("#1976D2","#388E3C", "#F57C00", "#D32F2F"))
```

From the visual we can see that no borrowers with a "good" credit rating
have any delinquencies and this is expected. Those with satisfactory make up
a small portion of delinquencies. There were a large number of borrowers with
very unsatisfactory credit ratings and current delinquencies in the lead up to
the GFC. Credit ratings in the very unsatisfactory bin appear to disappear from
2009 and this actually lowers the current delinquencies count significantly.

## Delinquencies by Employment Status and Credit Score
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, EmployDelCredit}
ggplot() + geom_bar(stat="identity", data = subset(loandf, 
                EmploymentStatus != ""), 
           aes(x=EmploymentStatus, y = CurrentDelinquencies, fill=CreditBins)) +
  scale_fill_manual(values = c("#1976D2","#388E3C", "#F57C00", "#D32F2F")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

The above bar graph is simple but it does elude to a number of things. The first
observation is that most groups contain some proportion of very unsatisfactory
credit score borrowers (with the exception of "Employed"). The "Employed"
category appears to have a higher proportion of borrowers with higher credit
scores than the rest of the groups. The highest proportion of "Very
Unsatisfactory credit score borrowers have come from the "Not Available"
employment status group. This seems somewhat suspicious, as if those with bad
credit scores are refraining from writing their employment status or lack there
of.

## Credit Binned Quarterly Delinquencies by Employment Status
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, EmployCredDelQuart}
ggplot() + geom_bar(stat="identity", data = subset(loandf, 
              loandf$EmploymentStatus != "" & !is.na(CreditBins)), 
           aes(x=LoanOriginationQuarter, y = CurrentDelinquencies, 
              fill=EmploymentStatus)) +
           theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  facet_wrap(~CreditBins, ncol = 2)
```

Described in final plot section

## Borrower APR vs Credit Score by Prosper Rating
<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, nonnull}
#creating a subset that removes NA values from variables to be used
filled <- subset(loandf, !is.na(DebtToIncomeRatio) & !is.na(BorrowerAPR) & 
                   !is.na(ProsperRating..numeric.))
```


```{r echo=FALSE, message=FALSE, warning=FALSE, CredRateAPR}
ggplot(aes(x = BorrowerAPR, y = as.integer(AverageCreditScore),
           color = factor(ProsperRating..numeric.), 
           group = factor(ProsperRating..numeric.)), 
           data = filled) +
  geom_jitter() +
    scale_color_manual(values = c("#FA4659", "#FC7055", "#FC9450", "#FAB448", 
                                  "#C0B743", "#7DB73F", "#00B33C")) +
  theme(legend.key.size = unit(0.9, "cm")) +
  theme_dark()
```


The above chart is quite detailed and there is a bit to digest. It shows the
relationship between average credit score and borrower APR for each of the
7 prosper ratings. Further discussion on this will be in the next section.

# Multivariate Analysis

### Were there any interesting or surprising interactions between features?

Yes, they will be described in the next section

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

Given my current time constraints I did not create any models with the dataset.
However I fully intend to at a later date. 

------

# Final Polished Plots and Summary

### Plot One
```{r  echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot(aes(x = IncomeRange, y = CurrentCreditLines), data = loandf) +
  geom_boxplot(lwd = 0.50, outlier.color = I("#C70039"), outlier.shape = 3) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Income Range and Credit Lines") +
  xlab("Income Range ($)") +
  ylab("Current Credit Lines") +
  scale_x_discrete(limits = c("Not employed", "$0", "$1-24,999", 
                              "$25,000-49,999", "$50,000-74,999", 
                              "$75,000-99,999", "$100,000+"))
```

### Description One

The above is a simple box plot shows a positive correlation between income and 
current credit lines. As we can see the median and positioning of the box 
increases as income does. The trend does not hold for those with $0 income. The
IQR also becomes larger with each income range jump. 

### Plot Two
```{r  echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot() + geom_bar(stat="identity", data = subset(loandf, 
              loandf$EmploymentStatus != "" & !is.na(CreditBins)), 
           aes(x=LoanOriginationQuarter, y = CurrentDelinquencies, 
              fill=EmploymentStatus)) +
           theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
      ggtitle("Credit Binned Quarterly Delinquencies by Employment Status") +
      xlab("Loan Origination Quarter") +
      ylab("Current Delinquencies") +
  facet_wrap(~CreditBins, ncol = 2)
```

### Description Two
This is actually quite a nice graph, albeit a little squashed when not in it's
own window. We can see that there are actually no delinquencies recorded for
those with a good credit score. The next level up "satisfactory" has limited
delinquenciy counts. When we get to the unsatisfactory level, we can see that
delinquency counts shoot up by quite a considerable amount. The very
unsatisfactory level had by far the highest counts pre GFC and after prosper
re-opened after it's hiatus, it appears that those with a credit score that low
have been unable to borrow on the platform. One interesting trend is the growing
delinquency counts among the individuals who have listed their employment status
as simply "Employed" which can have a very broad range of capacities. We can
also see that in each of the credit bins, there appears to be a spike in
delinquencies for those that are self-employed. Could these be the self employed
property investors/flippers of the sub prime crisis?

### Plot Three
```{r  echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(x = BorrowerAPR*100, y = as.integer(AverageCreditScore),
           color = factor(ProsperRating..numeric.), 
           group = factor(ProsperRating..numeric.)), 
           data = filled) +
  geom_jitter() +
  scale_color_manual(values = c("#FA4659", "#FC7055", "#FC9450", "#FAB448", 
                    "#C0B743", "#7DB73F", "#00B33C"), 
                    name = "Rating") +
  stat_smooth(method = 'lm', size = 1.3)+
  ggtitle("Borrower APR vs Credit Score by Prosper Rating") +
  xlab("BorrowerAPR (%)") +
  ylab("Average Credit Score") +
  theme(legend.key.size = unit(0.9, "cm")) +
  theme_dark()
```

### Description Three

There is quite a bit to digest in this scatter plot. The first observation is
that as credit score decreases, borrower APR increases (negative correlation).
The second observation is that the ratings prosper applies to their loans appear
to be fairly linked to the credit score and APR. We can see an almost rigid
colored stepping pattern as we move to lower credit scores and higher APR. I
have also added some lm lines so that we can see the trend lines for each of the
ratings. 

# Reflection

I am relatively new to R and as such this was a steep learning curve which I
hope to improve on as I learn more. The dataset was somewhat challenging, not
because of the length but because of the large amount of variables. I first had
to spend some time looking at the excel variable file and deciding what I would
examine before I even touched this report. A number of my charts were a little
disappointing as there was a lot of non-specific data. Examples of this include
the "Employed" category for employment and the lack of income range data
pre-GFC. 

Given the time constraints I had with balancing work with this project, I was
unable to construct any models. Future work would include analysing the
investing side of the data rather than just the debtor side. I would also like
to construct a model to predict defaults. 

# References

https://stackoverflow.com/questions/17496358/r-help-converting-factor-to-date

https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf